---
layout: topic
title: "Data exploration"
author: Tad, Priscilla San Juan
output: html_document
---

**Assigned Reading:**

> Zuur, A. F., E. N. Ieno, and C. S. Elphick. 2010. A protocol for data exploration to avoid common statistical problems. *Methods in Ecology and Evolution* **1**: 3-14.  [DOI: 10.1111/j.2041-210X.2009.00001.x](https://dx.doi.org/10.1111/j.2041-210X.2009.00001.x)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
```

### Key Points


### Analysis Example
```{r, include=FALSE, eval=TRUE}
# Load packages
library(phyloseq)
library(readr)
library(lattice)
library(qpcR)


# Load data
otu <- as.matrix((read_csv("~/Documents/R/Eco_stats_proj/Data/raw/OTU_tab_renamed.csv", col_names = TRUE)[,-1]))
tax <- as.matrix((read_csv("~/Documents/R/Eco_stats_proj/Data/raw/TAX_reorder_otu_name.csv", col_names = TRUE)[,-1]))

# Add in rownames
rownames(otu) <- paste0("OTU", 1:nrow(otu))
rownames(tax) <- paste0("OTU", 1:nrow(tax))

# Read file into phyloseq
OTU <- otu_table(otu, taxa_are_rows = TRUE)
TAX <- tax_table(tax)

# Combining OTU and TAX data
data <- phyloseq(OTU, TAX)

# Reading the sample data file
sample = sample_data(data.frame(read_csv("~/Documents/R/Eco_stats_proj/Data/raw/samplesinfo.csv")))

# Naming the the first rows using the first column
row.names(sample) = sample$ID

# Merging OTU/TAX file "Bac.data" with sampledata (includes MM and NEG)
merged <- merge_phyloseq(data, sample)

# Removing MM and NEG from merged_dataset
Real <- as.vector(data.frame(sample_data(merged))$Bird_species)
Real <- (!(Real%in%c("MM", "NEG", "RCRW")))
Real[is.na(Real)] = FALSE
merged_data = prune_samples(Real, merged)

# Remove OTUs with bad BLAST matches
kingdom <- as.vector(data.frame(tax_table(merged_data))$kingdom)
tax.keep <- kingdom=="Bacteria"
tax.keep[is.na(tax.keep)] = FALSE
merged_data <- prune_taxa(tax.keep,merged_data)

# Remove chloroplast (likely plant DNA)
class <- as.vector(data.frame(tax_table(merged_data))$class)
class <- (!(class%in%c("Chloroplast")))
class[is.na(class)] = FALSE
merged_data = prune_taxa(class, merged_data)

# Subset by species - BTSA
BTSA <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
BTSA <- BTSA%in%c("BTSA")
BTSA[is.na(BTSA)] = FALSE
data.BTSA <- prune_samples(BTSA, merged_data)

# Subset by species - CCRO
CCRO <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
CCRO <- CCRO%in%c("CCRO")
CCRO[is.na(CCRO)] = FALSE
data.CCRO <- prune_samples(CCRO, merged_data)

# Subset by species - OBNT
OBNT <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
OBNT <- OBNT%in%c("OBNT")
OBNT[is.na(OBNT)] = FALSE
data.OBNT <- prune_samples(OBNT, merged_data)

# Subset by species - RCWA
RCWA <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
RCWA <- RCWA%in%c("RCWA")
RCWA[is.na(RCWA)] = FALSE
data.RCWA <- prune_samples(RCWA, merged_data)

# Subset by species - SWTH
SWTH <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
SWTH <- SWTH%in%c("SWTH")
SWTH[is.na(SWTH)] = FALSE
data.SWTH <- prune_samples(SWTH, merged_data)

# Subset by species - YWAR
YWAR <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
YWAR <- YWAR%in%c("YWAR")
YWAR[is.na(YWAR)] = FALSE
data.YWAR <- prune_samples(YWAR, merged_data)

```

#### Checking for outliers

Zurr et al. recommends plotting your data using boxplots and dotcharts to detect outliers. Before removing suspected outliers, make sure they are actually outliers!


Since my data is multivariate, I used sequence count per sample for outlier detection in the following examples. 

**Boxplot**
```{r}
boxplot(sample_sums(merged_data),  ylab = "Sequences per sample")
```

**Cleveland dotchart**
```{r}
dotchart(sample_sums(merged_data), xlab = "Sequences per sample",
         ylab = "Samples", labels = F)
```
**Dotchart for multiple variables**

Since most of the variables in my dataset are categorical, here I used raw data (sequence count) subsetted by bird species.

```{r}
Z <- qpcR:::cbind.na(sample_sums(data.BTSA), sample_sums(data.CCRO), sample_sums(data.OBNT),
           sample_sums(data.RCWA), sample_sums(data.SWTH), sample_sums(data.YWAR))
colnames(Z) <- c("BTSA", "CCRO", "OBNT", "RCWA", "SWTH", "YWAR")
dotplot(as.matrix(Z), groups = FALSE,
        strip = strip.custom(bg = 'white',
        par.strip.text = list(cex = 0.8)),
        scales = list(x = list(relation = "free"),
                      y = list(relation = "free"),
                      draw = FALSE),
        col = 1, cex  = 0.5, pch = 16,
        xlab = "Number of sequences per sample",
        ylab = "Order of the data from text file")
```

#### Your data's distribution 

Some statistical tests assume normal distributions, making it important to check the shape of your data. Plot a histogram with all of your data and then groups of your data.

> "If you want to apply a statistical test to determine whether there is significant group separation in a discriminant analysis, however, normality of observations of a particular variable within each group is important" 

**Histogram**
```{r}
hist(sample_sums(merged_data), breaks = 100,col = "#a6bddb", xlab="Sequences per sample", main = ' ')
```


***


### Discussion Questions

**Q1:** When should you let go of an outlier?

**Q2:** How can you check for independence using multivariate data? 

> "Hence, it is important to check whether there is dependence in the raw data before doing the analysis, and also the residuals afterwards. These checks can be made by plotting the response variable vs. time or spatial coordinates."


**Q3:** _To transform or not to transform? That is the question._

> "There are three main reasons for a transformation: **to reduce the effect of outliers** (especially in covariates), **to stabilize the variance** and **to linearize relationships**. However, using more advanced techniques like GLS and GAMs, heterogenity and nonlinearity problems can be solved, making transformation less important."


**Q4:** What are some data exploration techniques you have used?



