---
layout: topic
title: "Data Exploration"
author: Tad
output: html_document
---

**Assigned Reading:**

> Zuur, A. F. 2010. *A protocol for data exploration to avoid common statistical problems.* Methods in Ecology and Evolution **1**: 3-14. DOI: [10.1111/j.2041-210X.2009.00001.x](http://dx.doi.org/10.1111/j.2041-210X.2009.00001.x)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
```

## Key Points

Test editing

The instructor will fill in a description of the topic here followed by a bulleted list describing key features of the statistical analysis method. For example:

* When to use this analysis
* When *not* to use this analysis

## Analysis Example

The student discussion leader will place their example code in this section, including text that describe the purpose of each code block.

First you wil probably need to read in data. If you have uploaded the data file on the course GitHub repository in the data folder then you should be able to read in the data as follows by changing the name of the file after `data/`:


```{r results = 'hide'}
# Read in a data file from the data folder on the BIO 202 GitHub repository
dat <- read.csv(file.path(repo_url, "data/macrolichen_traits_Coyle_2016.csv"))

# Or, read in the data file from a folder on your computer
# Remove # from the following 3 lines and change the file path assigned to my_dir
# my_dir <- "/Path/to/your/working_directory"
# setwd(my_dir)
# dat <- read.csv("lichen_traits_Coyle_2016.csv")
```


Next describe how you will manipulate the data in order to perform the anaylsis.

Include any relevant code below, making sure to include a `# comment` for each line.

```{r}
# Subset the data to foliose lichens
fol_dat <- subset(dat, Growth_form == 'foliose')

```


Describe the steps of the example analysis, including code blocks as needed. Describe the functions and packages used.

```{r chl_fig, echo = FALSE}
# echo = FALSE means that the results of the code will be shown without the code.
# chl_fig is a name that knit will use to name the figure generated by this code chunk

# Plot total chlorophyll vs. chlorophyll A:B ratio
plot(Tot_chl_DW ~ Chla2b, data = fol_dat, 
     las=1, 
     ylab = 'Tot. CHL',
     xlab = 'CHL A:B'
)

```

**Figure 1. Total chlorophyll versus ratio of chlorophyll A to B.** Looks like we have some data cleaning to do! 


## Discussion Questions

Include any questions or issues here that you want to discuss or ask about in class. After the class session, update this section with any important notes discussed.



