---
layout: topic
title: "Data exploration"
author: Tad, Priscilla San Juan
output: html_document
---

**Assigned Reading:**

> Zuur, A. F., E. N. Ieno, and C. S. Elphick. 2010. A protocol for data exploration to avoid common statistical problems. *Methods in Ecology and Evolution* **1**: 3-14.  [DOI: 10.1111/j.2041-210X.2009.00001.x](https://dx.doi.org/10.1111/j.2041-210X.2009.00001.x)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
```

### Key Points


### Analysis Example
```{r, include=FALSE, eval=TRUE}
library(phyloseq)
library(readr)
library(lattice)
library(qpcR)

otu <- as.matrix((read_csv("~/Documents/R/Eco_stats_proj/Data/raw/OTU_tab_renamed.csv", col_names = TRUE)[,-1]))
tax <- as.matrix((read_csv("~/Documents/R/Eco_stats_proj/Data/raw/TAX_reorder_otu_name.csv", col_names = TRUE)[,-1]))

# Add in rownames
rownames(otu) <- paste0("OTU", 1:nrow(otu))
rownames(tax) <- paste0("OTU", 1:nrow(tax))

# Read file into phyloseq
OTU <- otu_table(otu, taxa_are_rows = TRUE)
TAX <- tax_table(tax)

# Combining OTU and TAX data
data <- phyloseq(OTU, TAX)

# Reading the sample data file
sample = sample_data(data.frame(read_csv("~/Documents/R/Eco_stats_proj/Data/raw/samplesinfo.csv")))

# Naming the the first rows using the first column
row.names(sample) = sample$ID

# Merging OTU/TAX file "Bac.data" with sampledata (includes MM and NEG)
merged <- merge_phyloseq(data, sample)

# Removing MM and NEG from merged_dataset
Real <- as.vector(data.frame(sample_data(merged))$Bird_species)
Real <- (!(Real%in%c("MM", "NEG", "RCRW")))
Real[is.na(Real)] = FALSE
merged_data = prune_samples(Real, merged)

# Remove OTUs with bad BLAST matches
kingdom <- as.vector(data.frame(tax_table(merged_data))$kingdom)
tax.keep <- kingdom=="Bacteria"
tax.keep[is.na(tax.keep)] = FALSE
merged_data <- prune_taxa(tax.keep,merged_data)

# Remove chloroplast (likely plant DNA)
class <- as.vector(data.frame(tax_table(merged_data))$class)
class <- (!(class%in%c("Chloroplast")))
class[is.na(class)] = FALSE
merged_data = prune_taxa(class, merged_data)

# Subset by species - BTSA
BTSA <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
BTSA <- BTSA%in%c("BTSA")
BTSA[is.na(BTSA)] = FALSE
data.BTSA <- prune_samples(BTSA, merged_data)

# Subset by species - CCRO
CCRO <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
CCRO <- CCRO%in%c("CCRO")
CCRO[is.na(CCRO)] = FALSE
data.CCRO <- prune_samples(CCRO, merged_data)

# Subset by species - OBNT
OBNT <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
OBNT <- OBNT%in%c("OBNT")
OBNT[is.na(OBNT)] = FALSE
data.OBNT <- prune_samples(OBNT, merged_data)

# Subset by species - RCWA
RCWA <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
RCWA <- RCWA%in%c("RCWA")
RCWA[is.na(RCWA)] = FALSE
data.RCWA <- prune_samples(RCWA, merged_data)

# Subset by species - SWTH
SWTH <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
SWTH <- SWTH%in%c("SWTH")
SWTH[is.na(SWTH)] = FALSE
data.SWTH <- prune_samples(SWTH, merged_data)

# Subset by species - YWAR
YWAR <- as.vector(data.frame(sample_data(merged_data))$Bird_species)
YWAR <- YWAR%in%c("YWAR")
YWAR[is.na(YWAR)] = FALSE
data.YWAR <- prune_samples(YWAR, merged_data)

```

#### Checking for outliers

**Boxplot**
```{r}
boxplot(sample_sums(merged_data),  ylab = "Sequences per sample")
```

**Cleveland dotchart**
```{r}
dotchart(sample_sums(merged_data), xlab = "Sequences per sample",
         ylab = "Samples", labels = F)
```

```{r}
Z <- qpcR:::cbind.na(sample_sums(data.BTSA), sample_sums(data.CCRO), sample_sums(data.OBNT),
           sample_sums(data.RCWA), sample_sums(data.SWTH), sample_sums(data.YWAR))
colnames(Z) <- c("BTSA", "CCRO", "OBNT", "RCWA", "SWTH", "YWAR")
dotplot(as.matrix(Z), groups = FALSE,
        strip = strip.custom(bg = 'white',
        par.strip.text = list(cex = 0.8)),
        scales = list(x = list(relation = "free"),
                      y = list(relation = "free"),
                      draw = FALSE),
        col = 1, cex  = 0.5, pch = 16,
        xlab = "Number of sequences per sample",
        ylab = "Order of the data from text file")
```

```{r}
hist(sample_sums(merged_data), breaks = 100,col = "#a6bddb", main = "Histogram of sample sequencing depth")
```



```{r}
plot(table(round(otu_table(merged_data))),
    type = "h",
    xlim = c(0, 100),
    xlab = "Observed values", ylab = "Frequency")
```




### Discussion Questions

**Q1:** When should you let go of an outlier?

**Q2:** 

**Q3:** _To transform or not to transform? That is the question._

**Q4:** What are some data exploration techniques you have used?



