---
layout: topic
title: "GLM with zero-inflated data"
author: Tad; Nick Hendershot
output: html_document
---

**Assigned Reading:**

> Chapter 11 from: Zuur, A. F., Ieno, E. N., Walker, N., Saveliev, A. A. and Smith, G. M. 2009. *Mixed Effects Models and Extensions in Ecology with R.* Springer. [link](https://link-springer-com.stanford.idm.oclc.org/book/10.1007%2F978-0-387-87458-6)


```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/04-C/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
```

### Key Points



### Analysis Example

```{r, message=FALSE, warning=FALSE }
# Read in OBFL data

OBFL <- read.csv('./data/OBFL.csv', 
                 header = T)

############################################################
library(lattice)
library(MASS)
require(pscl) # alternatively can use package ZIM for zero-inflated models
library(lmtest)

```

Data includes counts of ochre-bellied flycatcher (OBFL) captures across deforestation and agricultural intensification gradient in southern Costa Rica. Specifically, sampling date, Counts (number of captures at site on a given day), Site Code and standardized mean canopy cover at the site (Canopy.std).

Here we will be looking at the relationship between number of captured individuals and canopy cover using a series of GLMs to deal with overdispersion. In reality we could use a GLMM with this data, but we will stick with GLMs for simplicity.


The steps and code use for these anlyses pull from heavily from:   
1. *Zuur, A. F., Ieno, E. N., Walker, N., Saveliev, A. A. and Smith, G. M. 2009. Mixed Effects Models and Extensions in Ecology with R.*    
2. *Zuur, A. F. and Ieno, E. N. 2016. Beginner's Guide to Zero-Inflated Models with R.*  

### Relationship between Counts and Canopy Cover
```{r pressure, echo=FALSE}
plot(Counts ~ Canopy.std,
     data = OBFL)
```

Let's do a quick check for Zero-Inflation in the data
```{r}
100*sum(OBFL$Counts == 0)/nrow(OBFL)
```
That's pretty high! ~ 40% of our data are zeros

While our data seems to be zero-inflated, this doesn't necessarily mean we need to use a zero-inflated model. In many cases, the covariates may predict the zeros under a Poisson or Negative Binomial model. So let's start with the simplest model, a Poisson GLM.

### Poisson GLM
$Counts_i \sim Poisson(\mu_i)$  
$E(Counts_i) = \mu_i$  
$var(Counts_i) = \mu_i$     
$log(\mu_i) = \alpha + \beta_1 * Canopy.std_i$  
```{r}
## Poisson GLM
M1 <- glm(Counts ~ Canopy.std,
          family = 'poisson',
          data = OBFL)

summary(M1)

## Check for over/underdispersion in the model
E2 <- resid(M1, type = "pearson")
N  <- nrow(OBFL)
p  <- length(coef(M1))   
sum(E2^2) / (N - p)


```
Looks like our model produces overdisperion. There are many reasons why this might be the case, but for now we are going to try to use a negative binomial GLM using the 'glm.nb' function in the 'MASS' package

### Negative Binomial GLM
$Counts_i \sim NB(\mu_i, theta)$  
$E(Counts_i) = \mu_i$  
$var(Counts_i) = (\mu_i + \mu_i^{2})/theta$   
$log(\mu_i) = \alpha + \beta_1 * Canopy.std_i$  
```{r}
M2 <- glm.nb(Counts ~ Canopy.std,
             data = OBFL)

summary(M2)



# Dispersion statistic
E2 <- resid(M2, type = "pearson")
N  <- nrow(OBFL)
p  <- length(coef(M2)) + 1  # '+1' is for variance parameter in NB
sum(E2^2) / (N - p)
```
Looks like the Negative Binomial GLM resulted in some minor underdispersion. In some cases, this might be OK. But in reality, we want to avoid both under- and overdispersion.  
Overdispersion can bias parameter estimates and produce false significant relationships.
On the otherhand, underdisperion can mask truly significant relationships. So let's try to avoid all of this.

### Zero-Inflated Poisson GLM
$Counts_i  \sim   ZIP(\mu_i, \pi_i)$      
$E(Counts_i) = \mu_i * (1-\pi_i)$  
$var(Counts_i) = (1-\pi_i) * (\mu_i + \pi_i * \mu_i^{2})$  
$log(\mu_i) = \beta_1 + \beta_2 * Canopy.std$  
$log(\pi_i) = \gamma_1 + \gamma_2 * Canopy.std$ 

In zero-inflated models, it is possible to choose different predictors for the counts and for the zero-inflation. You might expect different variables to be driving presence/absence vs. total number of individuals. We will keep it simple and use the same covariate in both parts.
```{r}
M3 <- zeroinfl(Counts ~ Canopy.std | ## Predictor for the Poisson process
                 Canopy.std, ## Predictor for the Bernoulli process;
               dist = 'poisson',
               data = OBFL)

summary(M3)

# Dispersion statistic
E2 <- resid(M3, type = "pearson")
N  <- nrow(OBFL)
p  <- length(coef(M3))  
sum(E2^2) / (N - p)
```
Still a some overdispersion in the ZIP. Let's try a ZINB

### Zero-Inflated Negative Binomial GLM
$Counts_i  \sim   ZINB(\mu_i, \pi_i)$  **_Not actually sure how to formulate this part for ZINB??_**    
$E(Counts_i) = \mu_i * (1-\pi_i)$  
$var(Counts_i) = (1-\pi_i) * (\mu_i + \pi_i^{2}/k) + \mu_i^{2} * (\pi_i^{2} + \pi_i)$  
$log(\mu_i) = \beta_1 + \beta_2 * Canopy.std$  
$log(\pi_i) = \gamma_1 + \gamma_2 * Canopy.std$ 

```{r}
M4 <- zeroinfl(Counts ~ Canopy.std |
                 Canopy.std,
               dist = 'negbin',
               data = OBFL)
summary(M4)

# Dispersion Statistic
E2 <- resid(M4, type = "pearson")
N  <- nrow(OBFL)
p  <- length(coef(M4)) + 1 # '+1' is due to theta
sum(E2^2) / (N - p)

```
This is really close to 1, which looks great.

We can compare the two zero-inflated models using the Likelihood ratio test
```{r}
lrtest(M3, M4)
```
Results show that the second model- the ZINB- is the best choice.
So let's stick with this model and plot the results


### Sketch fitted and predicted values
```{r, echo=FALSE}
MyData <- expand.grid(Canopy.std = seq(-1.31, 1.53, length = 25))

X.c <- model.matrix( ~ 1 + Canopy.std, data = MyData)
X.b <- model.matrix( ~ 1 + Canopy.std, data = MyData)

eta.mu <- X.c %*% coef(M4, model = 'count')
eta.Pi <- X.b %*% coef(M4, model = 'zero')


MyData$mu <- exp(eta.mu)
MyData$Pi <- exp(eta.Pi)/(1+exp(eta.Pi))
MyData$ExpY <- (1-MyData$Pi) * MyData$mu

par(mfrow = c(1,3), mar = c(5,5,2,2))

plot(x = OBFL$Canopy.std,
     y = OBFL$Counts,
     xlab = "\nMean Canopy\nCover (standardized)",
     ylab = "Zero inflated data",
     cex = 1,
     pch = 16,
     cex.lab = 1,
     cex.main = 1,
     main = "Count part")

lines(x = MyData$Canopy.std, 
      y = MyData$mu,
      lwd = 5)   

plot(x = MyData$Canopy.std,
     y = as.numeric(MyData$mu==0),
     xlab = "\nMean Canopy\nCover (standardized)",
     ylab = "Probability of false zero",
     cex = 1,
     pch = 16,
     cex.lab = 1,
     cex.main = 1,
     main = "Binary part",
     ylim = c(0,1))
      
lines(x = MyData$Canopy.std, 
      y = MyData$Pi,
      lwd = 5)         

plot(x = OBFL$Canopy.std,
     y = OBFL$Counts,
     xlab = "\nMean Canopy\nCover (standardized)",
     ylab = "Zero inflated data",
     cex = 1,
     pch = 16,
     main = "Expected values",
     cex.main = 1,
     cex.lab = 1)
lines(x = MyData$Canopy.std, 
      y = MyData$ExpY,
      lwd = 7)   

```


### Discussion Questions

1. When might you add differet covariates into the Bernoulli vs Poisson/NB portion of Zero-inflated models?  
2. How close to 1 do dispersion statistics need to be?
3. Anyone have experience with the two-step Zero-Altered models?






