---
title: "Visualizing spatial data"
author: "Tad & Tyler"
output:
  html_document: default
  pdf_document: default
layout: topic
---

**Assigned Reading:**

> Chapter 3 from: Bivand, R. S., Pebesma, E., Gomez-Rubio, V. 2013. *Applied Spatial Data Analysis with R*, 2nd ed. Springer. [Stanford Full Text](https://link-springer-com.stanford.idm.oclc.org/book/10.1007%2F978-1-4614-7618-4)

**Suggested Reading:**

> Chapters 2 and 8 from: Bivand, R. S., Pebesma, E., Gomez-Rubio, V. 2013. *Applied Spatial Data Analysis with R*, 2nd ed. Springer. [Stanford Full Text](https://link-springer-com.stanford.idm.oclc.org/book/10.1007%2F978-1-4614-7618-4)
>
> *Chapter 2 is an overview of how R handles spatial data. Chapter 8 is useful for mapping model-based spatial predictions.*



```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/08-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
```

### Key Points

#### Types of spatial data

There are two main ways to represent spatial data:

+ **raster**: a grid of values with a grain size and spatial extent
+ **vector**: a set of points, lines and polygons defined by their coordinates

The way spatial data are displayed depends on their projection or coordinate reference system (CRS). Some projections preserve distances between points whereas others presernce area or angles. *Know what the projection you are using does to distances before measuring distances between points using their coordinates.*

#### How R handles spatial data

The package `sp` has many of the base methods for handling spatial data in R. The package `rgdal` is needed to support different projections.

Objects in R that contain spatial information have a special class "Spatial". This means that they have information on the projection (stored as `proj4string`) and the broadest extent of the data (stored as a bounding rectangle/box `bbox`).

There are special classes for each type of spatial data:

+ `SpatialPoints` and `SpatialPointsDataFrame`
+ `SpatialLines` and `SpatialLinesDataFrame`
+ `SpatialPolygons` and `SpatialPolygonsDataFrame` (definitely read Chp 2 if you need to work with polygon data)
+ `SpatialGrid` and `SpatialGridDataFrame` or `SpatialPixels` and `SpatialPixelsDataFrame` (use the `raster` package if you are mostly working with gridded data- it stores information much more efficiently and is easier to use).

Use the `CRS()` function to add a projection to data to turn it into spatial data. `CRS` accepts text formatted according to the proj.4 framework, which looks like:

`CRS("+proj=longlat +ellps=WGS84")`

This tells R that the coordinates in the associated data are in longitude and latitude and that the model of the shape of the earth that should be used (WGS84). [This tutorial](http://rspatial.org/spatial/rst/6-crs.html) gives an explanation of coordinate reference systems in R and its parent website is a great resources for spatial analysis in R.

Other links:

+ [Here](http://proj4.org/projections/index.html) is a list of how to specify the different projections in the proj.4 framework.
+ [Here](http://proj4.org/parameters.html#parameter-list) is a list of how to specify other parameters, like the center of the projection and the units.

#### Visualizing spatial data in R

**Base graphics** 

Use: `plot()` and `image()`

Pros:

  + build layer by layer with separate function calls
  + same basic functions used to plot regular data

Cons:

  + duplicated code when plotting the same map for multiple variables
  + manually map variables to visual attributes (must create your own scales)
  + manually make legends


**Grid graphics (e.g. `lattice`)**

`spplot()` is based on `xyplot()` for point data and `levelplot()` for grid data.

Pros:

  + conditional plots (same code makes multiple maps with different variables)
  + automatic scaling of variables mapped to visual attributes
  + automatic legends

Cons:

  + dense and cryptic documentation makes it hard to learn how to customize
  
**ggplot2**

Use: `ggplot()`

Pros:

  + no special handling of spatial data once spatial data is converted to a dataframe
  + conditional plots (same code makes multiple maps with different variables)
  + automatic scaling of variables mapped to visual attributes
  + automatic legends

Cons:

  + `ggplot` doesn't work automatically on SpatialPointsDataFrames or gridded data.
  + `geom_grid` and `geom_tile` are for gridded data but may be inefficient for very large grids because data must be supplied in long form rather than a matrix of values.


#### Colors

Useful functions:

+ `colorRampPalette()` makes a color ramp between the indicated colors with a given number of colors
+ `classIntervals()` creates set of intervals between two numbers with breaks between intervals defined in different ways.


### Analysis Example

**Traditional plot system (base graphics)**

mesuse is a dataframe with 155 obs. of 14 variables. First two columns are lat/long identifying sampling sites. The remaining collumns are environmental variables collected at each point. Each point represents as sampling site in the flood plain of the river Meuse in the Netherlands. Researchers were interested in topsoil heavy metal concentrations.

## Plotting geometry

# Points
```{r}
library(sp) 
data(meuse) 
coordinates(meuse) <- c("x", "y") # tells r to use x and y collumns as the lat/long 
plot(meuse) # by default plots the points
```

# Lines

```{r}
cc <- coordinates(meuse) 
m.sl <- SpatialLines(list(Lines(list(Line(cc)), "line1"))) 
plot(m.sl) # plots lines connecting all the points
```

# Polygons

```{r}
data(meuse.riv)  # lat/longs outlining bank or river
meuse.lst <- list(Polygons(list(Polygon(meuse.riv)), "meuse.riv")) # not sure what this is doing?
meuse.pol <- SpatialPolygons(meuse.lst) # converts list to SpatialPolygon, which can be plotted
plot(meuse.pol, col = "grey") 

str(meuse.lst)
str(meuse.pol)

str(meuse)
```

# Grid

```{r}
data(meuse.grid) # from GIS data?
coordinates(meuse.grid) <- c("x", "y") 
meuse.grid <- as(meuse.grid, "SpatialPixels") 
image(meuse.grid, col = "grey") 

```

# Combining multiple elements

You can combine multiple layers by using the add = TRUE argument

```{r}
image(meuse.grid, col = "yellow") 
plot(meuse.pol, col = "blue", add = TRUE) 
plot(meuse, pch = 1, cex = 0.5, add = TRUE) # change shape and size of points
# see table 3.2 for helpful arguments for customizing plots

```

## Axes and layout options

```{r}
layout(matrix(c(1, 2), 1, 2)) 
plot(meuse.pol, axes = TRUE) # default axes
plot(meuse.pol, axes = FALSE) # with custom axes
axis(1, at = c(178000 + 0:2 * 2000), cex.axis = 0.7) 
axis(2, at = c(326000 + 0:3 * 4000), cex.axis = 0.7) 
box()
```

# Plot without axes

```{r}
oldpar = par(no.readonly = TRUE) 
layout(matrix(c(1, 2), 1, 2))
plot(meuse, axes = TRUE, cex = 0.6) 
plot(meuse.pol, add = TRUE) 
title("Sample locations") 
par(mar = c(0, 0, 0, 0) + 0.1)
plot(meuse, axes = FALSE, cex = 0.6) 
plot(meuse.pol, add = TRUE) 
box() 
par(oldpar)
```


# Adding degrees and gridlines

```{r}
# Install and load packages

#install.packages("maptools")
library(maptools) 
library(maps) 
#install.packages("rgdal")
library(rgdal) 

```
```{r}
wrld <- map("world", interior = FALSE, xlim = c(-179, + 179), ylim = c(-89, 89), 
            plot = FALSE) 
wrld_p <- pruneMap(wrld, xlim = c(-179, 179)) 
llCRS <- CRS("+proj=longlat +ellps=WGS84") 
wrld_sp <- map2SpatialLines(wrld_p, proj4string = llCRS) 
prj_new <- CRS("+proj=moll") 

wrld_proj <- spTransform(wrld_sp, prj_new) 
wrld_grd <- gridlines(wrld_sp, easts = c(-179, seq(-150, + 150, 50), 179.5), 
                      norths = seq(-75, 75, 15), ndiscr = 100) 
wrld_grd_proj <- spTransform(wrld_grd, prj_new) 
at_sp <- gridat(wrld_sp, easts = 0, norths = seq(-75, + 75, 15), offset = 0.3) 
at_proj <- spTransform(at_sp, prj_new) 
plot(wrld_proj, col = "grey60") 
plot(wrld_grd_proj, add = TRUE, lty = 3, col = "grey70") 
text(coordinates(at_proj), pos = at_proj$pos, offset = at_proj$offset, 
     labels = parse(text = as.character(at_proj$labels)), cex = 0.6)
```


## Plotting attributes and map legends

Up to now we have only plotted the geometry or topology of the spatial objects. If in addition we want to show feature characteristics or attributes of the objects, we need to use type, size, or colour of the symbols, lines, or polygons.

```{r}
#install.packages("gstat")
library(gstat) 
```


```{r}
idw.out <- gstat::idw(zinc ~ 1, meuse, meuse.grid, idp = 2.5) # gstat package used to create objects that hold info used for geostatistical prediction. idw is inverse distance weighted interpolation.
grays = gray.colors(4, 0.55, 0.95) # defines color scheme
image(idw.out, col = grays, breaks = c(100, 200, 400, 800, 1800))  
plot(meuse.pol, add = TRUE) 
plot(meuse, pch = 1, cex = sqrt(meuse$zinc)/40, add = TRUE) 
legVals <- c(100, 200, 500, 1000, 2000) 
legend("bottomleft", legend = legVals, pch = 1, pt.cex = sqrt(legVals)/40, bty = "n", title = "measured") 
legend("topleft", legend = c("100-200", "200-400", "400-800", "800-1800"), fill = grays, bty = "n", title = "interpolated")


```


**Tellis/lattice plots with spplot**

spplot provides plotting of spatial data with attributes through the Trellis graphics system. Good for making conditional plots.

# Plotting points, polygons, lines, and grids

```{r}
library(lattice)
spplot(meuse, c("cadmium", "zinc", "copper", "lead")) # easy way to plot side by side maps
```

```{r}
library(maptools) 
data(meuse.grid) 
coordinates(meuse.grid) <- c("x", "y") 
meuse.grid <- as(meuse.grid, "SpatialPixelsDataFrame") 
im <- as.image.SpatialGridDataFrame(meuse.grid["dist"]) 
cl <- ContourLines2SLDF(contourLines(im)) 
spplot(cl)

```


# Adding reference and layout elements to plots

```{r}
river <- list("sp.polygons", meuse.pol) 
north <- list("SpatialPolygonsRescale", layout.north.arrow(), offset = c(178750, 332500), scale = 400) 
scale <- list("SpatialPolygonsRescale", layout.scale.bar(), offset = c(180200, 329800), scale = 1000, fill = c("transparent", "black"))
txt1 <- list("sp.text", c(180200, 329950), "0") 
txt2 <- list("sp.text", c(181200, 329950), "1 km") 
pts <- list("sp.points", meuse, pch = 3, col = "black") 
meuse.layout <- list(river, north, scale, txt1, txt2, pts)
zn <- meuse["zinc"]
spplot(zn, sp.layout = meuse.layout)

```

** Other options: ggplot2 and latticeExtra**

```{r}
library(ggplot2) 
methods(fortify) # use fortify function to convert spatial data to a dataframe which ggplot can read
```


However, doesn't work for SpatialPointsDataFrame such as meuse data set, so we have to do it manually.

```{r}
m = as(meuse, "data.frame") 
ggplot(m, aes(x, y)) + geom_point() + coord_equal()
```


# latticeExtra

```{r}
library(latticeExtra) 
p = spplot(meuse["zinc"]) 
m = SpatialPolygonsDataFrame(meuse.pol, data.frame(col = 1), match.ID = FALSE) 
l = spplot(m) 
l+p 
p+l
```


### Additional resources

# About how R handles spatial data

http://rspatial.org/spatial/rst/1-introduction.html 

# About making maps in ggplot

http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html

https://timogrossenbacher.ch/2016/12/beautiful-thematic-maps-with-ggplot2-only/


### Discussion Questions

*1)* Which of the three approaches (base plot, lattice, ggplot2) do you prefer? 

*2)* What kinds of spatial data do you have and what kinds of visuals have you used?

*3)* Have you used any other usefull packages for vizualizing spatial data?

*4)* Pros/cons of using R vs. GIS or other platforms for data visualization?


